---
title: "GxE_final_project"
author: "Vasiliy Zubarev"
date: "10 05 2023"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(dplyr)
library(tidyr)
library(data.table)
library(ggplot2)
library(stats)
#MLMs
library(lme4)
#heatmaps of correlations
library(pheatmap)
#PCA visualization
library(FactoMineR)
library(factoextra)

#Finlay-Winkinson regression
#library(devtools)
#install_github("lian0090/FW")
library(FW)
```

```{r}
all_pheno<-read.csv('SupplTableS1_phenotypes_rawData.txt', header=TRUE, sep=' ')
head(all_pheno)
```

##Numbers of observations

```{r}
all_pheno %>% count(Environment)
```

Explore observation counts per line and environment:

```{r}
GxE_counts<-all_pheno %>% count(Genotype, Environment)
ggplot(GxE_counts, aes(x=Environment, y=Genotype, fill=n))+geom_tile()+theme_classic()+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.text.y = element_blank())+ggtitle("Number of observations per genotype and environment")
```

The dataset is highly unbalanced. Most lines have 4 observations in several environments, but some lines have another number of observations.

##Percentage of NAs

1) Simply by each row

```{r}
colMeans(is.na(all_pheno))
```

We can see that some traits have a lot of NAs.

2) By traits and environment groups:

```{r}
nas<-as.data.frame(is.na(all_pheno[,c(9:17)]))
nas$Environment<-all_pheno$Environment
setDT(nas)
nas_summary<-nas[, lapply(.SD, mean), by=Environment]
nas_summary<-gather(nas_summary, key="parameter", value="value", 2:10)
ggplot(nas_summary, aes(x=parameter, y=Environment, fill=value))+geom_tile()+theme_classic()+ggtitle("NA fraction per place, year and parameter")
```

## Trait distributions

```{r}
par(mfrow=c(3,3))
hist(all_pheno$EV_V4)
hist(all_pheno$EV_V6)
hist(all_pheno$PH_V4)
hist(all_pheno$PH_V6)
hist(all_pheno$PH_final)
hist(all_pheno$LO)
hist(all_pheno$TILL)
hist(all_pheno$FF)
hist(all_pheno$MF)
```

## Correlation analysis

To understand dependencies between phenotypic variables, we use Pearson correlations.

```{r}
phenotypes<-all_pheno[c(9:17)]

#get rid rows with NAs
phenotypes<-phenotypes[complete.cases(phenotypes),]

#compute correlations
cor(phenotypes)
cor_pheno<-data.frame(cor(phenotypes))

#plot correlation heatmap
pheatmap(cor_pheno)
```

We see 2 obvious groups of traits. 

A) EV_V4, EV_V6, PH_V4, PH_V6 positively correlates with p=0.7...0.85.

B) MF and FF positively correlates with p=0.78.

Between these groups there are negative correlations with p=-0.35...-0.5

## Principal Components Analysis

To understand which traits are the most discriminative among all, we use PCA.

```{r}
pca<-PCA(phenotypes)
pca$var
```

```{r}
#par(mfrow=c(1,2))
fviz_pca_biplot(pca,
             geom.ind = "point", 
             col.ind="blue",
             col.var="black",
             addEllipses = TRUE,
             alpha.ind = 0.15, 
             legend.title = "Groups"
)
fviz_pca_biplot(pca,
            axes=c(2,3),
             geom.ind = "point", 
             col.ind="blue",
             col.var="black",
             addEllipses = TRUE,
             alpha.ind = 0.15, 
             legend.title = "Groups"
)
```

```{r}

```

## Introducing mixed linear models to find the best lines for further breeding

For EV_V4, EV_V6, PH_final, PH_V4, PH_V6 we have measures in all environments, FF - in all but one. Let's start with these traits. Also, some environments does not have all lines measured. Let's start with environments that have all lines.

Correlated variables EV_V4, EV_V6, PH_V4, PH_V6 will be scaled and considered together as a mean of scaled variables.

Firstly drop GOL and TOM sites due to incomplete lines set. Secondly drop lines that are not measured in other sites.

```{r}
EV_V4_morelines<-all_pheno[c(1:12)] #all predictors and EV_V4 trait
head(EV_V4_morelines)

#drop environments having a lot of missing data
EV_V4_morelines_filt<-EV_V4_morelines[! EV_V4_morelines$Environment %in% c("2017.GOL","2017.TOM","2018.GOL","2018.TOM"),]

n_genot<-EV_V4_morelines_filt$Genotype %>% unique() %>% length()
cat(n_genot, " genotypes prior to the filtering\n")

#drop rows with NAs
EV_V4_morelines_filt<-EV_V4_morelines_filt[complete.cases(EV_V4_morelines_filt),]

#filter only lines that are measured in all remaining envs/years
obs_count<-EV_V4_morelines_filt %>% group_by(Genotype, Environment) %>% count() %>% spread(key=Environment, value=n)
obs_count<-obs_count[complete.cases(obs_count),]
obs_count
complete_genot<-obs_count$Genotype
EV_V4_morelines_filt<-EV_V4_morelines_filt[EV_V4_morelines_filt$Genotype %in% complete_genot,]

#drop rows with NAs again
EV_V4_morelines_filt<-EV_V4_morelines_filt[complete.cases(EV_V4_morelines_filt),]

cat(length(complete_genot), " genotypes after the filtering")
```

Scaling and computing mean

```{r}
#EV_V4_morelines_filt$EV_V4<-scale()
EV_V4_morelines_filt$EV_V4<-scale(EV_V4_morelines_filt$EV_V4)
EV_V4_morelines_filt$EV_V6<-scale(EV_V4_morelines_filt$EV_V6)
EV_V4_morelines_filt$PH_V4<-scale(EV_V4_morelines_filt$PH_V4)
EV_V4_morelines_filt$PH_V6<-scale(EV_V4_morelines_filt$PH_V6)
EV_V4_morelines_filt$EV<-(EV_V4_morelines_filt$EV_V4+EV_V4_morelines_filt$EV_V6+EV_V4_morelines_filt$PH_V4+EV_V4_morelines_filt$PH_V6)/4
head(EV_V4_morelines_filt)
hist(EV_V4_morelines_filt$EV)
```

```{r}
write.csv(EV_V4_morelines_filt[c(1:8,13)], "earlyvigour_filtered_7envs.csv", row.names=FALSE, quote=FALSE )
```


```{r}
sum(is.na(EV_V4_morelines_filt))
tail(EV_V4_morelines_filt)
```

```{r}
mixed.lmer <- lmer(EV ~ Environment+(1|Genotype)+(1|Block), data = EV_V4_morelines_filt)
```

```{r}
plot(mixed.lmer)
qqnorm(resid(mixed.lmer))
qqline(resid(mixed.lmer))
```

Define a function to get BLUPs (from Alexey Zamalutdinov)

```{r}
pev_rel_estimation <- function(model) {
  # Using Henderson's equations
  
  X <- getME(model,'X')
  Z <- getME(model,'Z')
  Y <- getME(model,'y')
  
  # model w/o genetic relationships
  varianceestimates <- as.data.frame(VarCorr(model))[,4]
  se <- varianceestimates[2]
  su <- varianceestimates[1]
  
  lambda <- se/su
  Iu <- diag(length(levels(as.factor(EV_V4_morelines_filt$Genotype)))+length(levels(as.factor(EV_V4_morelines_filt$Block))))
  #length(levels(as.factor(EV_V4_morelines_filt$Genotype)))+length(levels(as.factor(EV_V4_morelines_filt$Block)))
  
  XpX <- crossprod(X)
  XpZ <- crossprod(X, Z)
  ZpX <- crossprod(Z, X)
  ZpZ <- crossprod(Z)
  XpY <- crossprod(X, Y)
  ZpY <- crossprod(Z, Y)
  

  ## LHS
  LHS <- rbind(cbind(XpX, XpZ),
               cbind(ZpX, ZpZ + Iu * lambda))
  #print(LHS)
  
  ## RHS
  RHS <- rbind(XpY, 
               ZpY)
 
  # Inverse of LHS
  InvLHS <- round(solve(LHS), digit=3) ; size <- dim(InvLHS)[[1]]
  #print(InvLHS)
  cat("size: ",size)
  PEVpre <- diag(InvLHS)[ 8 : size]
  #print(PEVpre)
  #' and then corresponding values:
  PEV <- PEVpre * se
  SEP <- sqrt(PEV)
  REL <- 1 - PEV / su
  
  solutions <- as.data.frame(ranef(model)) 
  cat("solutions len ",length(solutions[, 3]),"\n")
  cat("SEP len ",length(SEP),"\n")
  cat("PEV len ",length(PEV),"\n")
  return(data.frame(variety_name = solutions[, 3],
                    BLUP = solutions[, 4],
                    PEV = PEV,
                    SEP = SEP,
                    REL = REL))
}
```

```{r}
answer<-pev_rel_estimation(mixed.lmer)
```

```{r}
genot_EV_BLUPs<-answer[c((1717-696):1717),c(1,2)]
genot_EV_BLUPs<-genot_EV_BLUPs[order(-genot_EV_BLUPs$BLUP),]
(topten_list<-genot_EV_BLUPs[c(1:10),1])
topten_df<-all_pheno[all_pheno$Genotype %in% topten_list,]
topten_list
#drop rows with NAs
#topten_df<-topten_df[complete.cases(topten_df),]
#write.csv(topten_df, "./earlyvigor_top10_allpheno.csv", row.names=FALSE, quote=FALSE)
```

## Finley-Wilkinson

```{r}
Indexes <- EV_V4_morelines_filt %>% group_by(Environment) %>% summarise(Index = mean(EV))
Indexes$Index <- Indexes$Index - mean(EV_V4_morelines_filt$EV) ## do not forget
Indexes
EV_V4_morelines_filt <- left_join(EV_V4_morelines_filt, Indexes)
head(EV_V4_morelines_filt)
```

```{r}
JointReg <- lmer(EV ~ Environment + Genotype + Genotype:Index + (1|Block), data = EV_V4_morelines_filt)
```

```{r}
plot(JointReg)
qqnorm(resid(JointReg))
qqline(resid(JointReg))
```

```{r}
toto <- data.frame(EV_V4_morelines_filt, predicted = predict(JointReg))
head(toto)
```

```{r}
toto %>% filter(Genotype %in% topten_list) %>% ggplot(aes(x = Index, y = predicted, group = Genotype,
colour = Genotype)) + geom_point() + geom_smooth(method='lm', formula='y~x', se=FALSE) + geom_abline(slope = 1,
intercept = mean(EV_V4_morelines_filt$EV), lty = 2) + geom_vline(xintercept = 0, lty = 2) + theme(plot.background = element_rect(fill = "aliceblue"),
plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))
```

## Analyze flowering time

For a first time, by simple mixed-effects model with no GxE

```{r}
Flo<-all_pheno[,c(1:8,16,17)]
Flo$MF<-scale(Flo$MF)
Flo$FF<-scale(Flo$FF)
Flo$FT<-rowMeans(Flo[,9:10], na.rm=TRUE)
#drop rows with NAs again
Flo<-Flo[complete.cases(Flo$FT),]
head(Flo)
```

```{r}
lmer_flo <- lmer(FT ~ Environment+(1|Genotype)+(1|Block), data = Flo)
```

```{r}
plot(lmer_flo)
qqnorm(resid(lmer_flo))
qqline(resid(lmer_flo))
```

```{r}
pev_rel_estimation <- function(model) {
  # Using Henderson's equations
  
  X <- getME(model,'X')
  Z <- getME(model,'Z')
  Y <- getME(model,'y')
  
  # model w/o genetic relationships
  varianceestimates <- as.data.frame(VarCorr(model))[,4]
  se <- varianceestimates[2]
  su <- varianceestimates[1]
  
  lambda <- se/su
  Iu <- diag(length(levels(as.factor(Flo$Genotype)))+length(levels(as.factor(Flo$Block))))
  #length(levels(as.factor(EV_V4_morelines_filt$Genotype)))+length(levels(as.factor(EV_V4_morelines_filt$Block)))
  
  XpX <- crossprod(X)
  XpZ <- crossprod(X, Z)
  ZpX <- crossprod(Z, X)
  ZpZ <- crossprod(Z)
  XpY <- crossprod(X, Y)
  ZpY <- crossprod(Z, Y)
  

  ## LHS
  LHS <- rbind(cbind(XpX, XpZ),
               cbind(ZpX, ZpZ + Iu * lambda))
  #print(LHS)
  
  ## RHS
  RHS <- rbind(XpY, 
               ZpY)
 
  # Inverse of LHS
  InvLHS <- round(solve(LHS), digit=3) ; size <- dim(InvLHS)[[1]]
  #print(InvLHS)
  cat("size: ",size)
  PEVpre <- diag(InvLHS)[ 11 : size]
  #print(PEVpre)
  #' and then corresponding values:
  PEV <- PEVpre * se
  SEP <- sqrt(PEV)
  REL <- 1 - PEV / su
  
  solutions <- as.data.frame(ranef(model)) 
  cat("solutions len ",length(solutions[, 3]),"\n")
  cat("SEP len ",length(SEP),"\n")
  cat("PEV len ",length(PEV),"\n")
  return(data.frame(variety_name = solutions[, 3],
                    BLUP = solutions[, 4],
                    PEV = PEV,
                    SEP = SEP,
                    REL = REL))
}
```

```{r}
answer_Flo<-pev_rel_estimation(lmer_flo)
```

```{r}
(genot_FT_BLUPs<-answer[c(1021:1973),c(1,2)])
(genot_FT_BLUPs<-genot_FT_BLUPs[order(genot_FT_BLUPs$BLUP),])
(topten_list_Flo<-genot_FT_BLUPs[c(1:10),1])
topten_df_Flo<-all_pheno[all_pheno$Genotype %in% topten_list,]
topten_list_Flo
#drop rows with NAs
#topten_df<-topten_df[complete.cases(topten_df),]
#write.csv(topten_df, "./earlyvigor_top10_allpheno.csv", row.names=FALSE, quote=FALSE)
```


```{r}
```


With a FW regression

```{r}
Indexes <- Flo %>% group_by(Environment) %>% summarise(Index = mean(FT))
Indexes$Index <- Indexes$Index - mean(Flo$FT) ## do not forget
Indexes
Flo <- left_join(Flo, Indexes)
head(Flo)
```

```{r}
JointRegFlo <- lmer(FT ~ Environment + Genotype + Genotype:Index + (1|Block), data = Flo)
```

```{r}
#summary(JointRegFlo)
Flo_geneff<-data.frame(fixef(JointRegFlo))
Flo_geneff$Genotype<-rownames(Flo_geneff)
Flo_geneff<-data.frame(Flo_geneff[c(11:nrow(Flo_geneff)),])
(Flo_geneff<-Flo_geneff[order(Flo_geneff$fixef.JointRegFlo.),])
```

```{r}
plot(JointRegFlo)
qqnorm(resid(JointRegFlo))
qqline(resid(JointRegFlo))
```

```{r}

```


```{r}
toto_Flo <- data.frame(Flo, predicted = predict(JointRegFlo))
toto_Flo %>% filter(Genotype %in% topten_list_Flo)
```

```{r}
toto_Flo %>% filter(Genotype %in% c("DH_PE0539", "DH_LL0018", "DH_LL0018", "DH_PE0452", "DH_PE0025", "DH_KE0281", "DH_KE0239", "DH_PE0423", "DH_KE0239", "DH_KE0188", "DH_PE0060", "DH_PE0506")) %>% ggplot(aes(x = Index, y = predicted, group = Genotype,
colour = Genotype)) + geom_point() + geom_smooth(method='lm', formula='y~x', se=FALSE) + geom_abline(slope = 1,
intercept = mean(Flo$FT), lty = 2) + geom_vline(xintercept = 0, lty = 2) + theme(plot.background = element_rect(fill = "aliceblue"),
plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))
```

## Analyzing lodging score

```{r}
Lod<-all_pheno[,c(1:8,14)]
#drop rows with NAs again
Lod<-Lod[complete.cases(Lod$LO),]
head(Lod)
```

```{r}
#lmer_lod<-lmer(LO ~ Environment + Genotype + Genotype:Index+(1|Block), data = Lod)
```

```{r}
#lmer_without_int_lod<-lmer(LO ~ Environment+(1|Genotype)+(1|Block), data = Lod)
```


```{r}
#plot(lmer_without_int_lod)
#qqnorm(resid(lmer_without_int_lod))
#qqline(resid(lmer_without_int_lod))
```

```{r}
#plot(lmer_lod)
#qqnorm(resid(lmer_lod))
#qqline(resid(lmer_lod))
```

```{r}
#summary(JointRegFlo)
#Lod_geneff<-data.frame(fixef(lmer_lod))
#Lod_geneff$Genotype<-rownames(Lod_geneff)
#Lod_geneff<-data.frame(Lod_geneff[c(5:nrow(Lod_geneff)),])
#(Lod_geneff<-Lod_geneff[order(Lod_geneff$fixef.lmer_lod.),])
```

```{r}
#toto_Lod %>% filter(Genotype %in% c("DH_PE0053", "DH_PE0279", "DH_PE0386", #"DH_PE0039", "DH_KE0474", "DH_KE0631", "DH_PE0320", "DH_PE0113", #"DH_PE0007")) %>% ggplot(aes(x = Index, y = predicted, group = Genotype,
#colour = Genotype)) + geom_point() + geom_smooth(method='lm', #formula='y~x', se=FALSE) + geom_abline(slope = 1,
#intercept = mean(Lod$LO), lty = 2) + geom_vline(xintercept = 0, lty = 2) + #theme(plot.background = element_rect(fill = "aliceblue"),
#plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))
```

```{r}
#toto_Lod <- data.frame(Lod, predicted = predict(lmer_lod))
```

```{r}
#lm1 <- FW(y = METmaize$yield, VAR = METmaize$Genotype, ENV = #METmaize$Environ,
#method = "OLS")
```

```{r}
```



Predict EV value from the model, excluding block term

```{r}
#form<-EV ~ Environment+(1|Genotype)+(1|Block)
#(bar.f <- findbars(form)) # list with 3 terms
#mf <- model.frame(subbars(form),data=EV_V4_morelines_filt)
#rt <- mkReTrms(bar.f,mf)
#names(rt)
```

```{r}
#x<-EV_V4_morelines_filt[c(1,2), c('Genotype', 'Environment', 'Block')]
#x
#predict(mixed.lmer, newdata=x, re.form=rt, allow.new.levels = TRUE)
```


```{r}
#EV_V4_morelines_model<-lmer()
```

```{r}
#n=5
#cat("n = ", n)
```






Find the subset of lines that are measured in completely all environments

```{r}
filtered<-all_pheno #all predictors and EV_V4 trait

#drop environments having a lot of missing data
EV_V4_morelines_filt<-EV_V4_morelines[! EV_V4_morelines$Environment %in% c("2017.GOL","2017.TOM","2018.GOL","2018.TOM"),]

n_genot<-filtered$Genotype %>% unique() %>% length()
cat(n_genot, " genotypes prior to the filtering\n")

#filter only lines that are measured in all remaining envs/years
obs_count<-filtered %>% group_by(Genotype, Environment) %>% count() %>% spread(key=Environment, value=n)
obs_count<-obs_count[complete.cases(obs_count),]
obs_count
complete_genot<-obs_count$Genotype
filtered<-filtered[filtered$Genotype %in% complete_genot,]

#drop rows with NAs again
#EV_V4_morelines_filt<-EV_V4_morelines_filt[complete.cases(EV_V4_morelines_filt),]



cat(length(complete_genot), " genotypes after the filtering")
```

```{r}
GxE_filtcounts<-filtered %>% count(Genotype, Environment)
ggplot(GxE_filtcounts, aes(x=Environment, y=Genotype, fill=n))+geom_tile()+theme_classic()+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.text.y = element_blank())+ggtitle("Number of observations per genotype and environment")
```

```{r}
nas<-as.data.frame(is.na(filtered[,c(9:17)]))
nas$Environment<-filtered$Environment
setDT(nas)
nas_summary<-nas[, lapply(.SD, mean), by=Environment]
nas_summary<-gather(nas_summary, key="parameter", value="value", 2:10)
ggplot(nas_summary, aes(x=parameter, y=Environment, fill=value))+geom_tile()+theme_classic()+ggtitle("NA fraction per place, year and parameter")
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
#mean(c(1,NA,3), )
```

